NumPy---Array Computation for Python

### 2020-01-06 body.tex
Added notes and small edits

NumPy is both a cause and an effect of a striking shift in the nature of
scientific computing.  The traditional model of scientific software has been of
the user as {\it consumer}.  Writing software is a task for specialists and
professionals, supported by income from selling the software to scientists to
use.  The new model is of the user as {\it owner}.  The Python language, with
the NumPy library, provides a framework that scientists can consume, but has
also been successful in encouraging scientists to develop as programmers, and
to write code for the library, and the system of packages around it.  This has
fostered a new culture of contribution and collaboration in scientific
computing that resulted in a virtuous cycle of increasing use, and increasing
development.

@@

The first incarnation of NumPy was a package called Numeric, written in 1995
by Jim Hugunin, then completing his master's thesis at MIT on
superconductor-semiconductor junctions.  Hugunin based his package on previous
work by Jim Fulton, then working at the US Geological Survey, and acknowledges
many others in his initial report of the package \cite{Hugunin-whitepaper}.
Numeric provided an array object in Python, written in C, and linking to
standard fast implementations of linear algebra.

As Hugunin points out, at the time he wrote Numeric, commercial languages for
interactive computing with numerical arrays were already well-established,
including Matlab and IDL.  At the time, it must have seemed a safe bet that
Numeric would not succeed in this market.  The commercial languages were
developed and maintained by large numbers of full-time professional programmers
and supported by formal quality assurance, release managers and writers of
scientific documentation.  In contrast, Numeric, and later, NumPy, would
continue to be developed almost entirely by scientists, including graduate
students and faculty, without specific funding, and often to the detriment of
the scientists' careers.

@@

Around 1998, the Space Telescope Science Institute (STScI) software group began
to use Python heavily, and, around 2000, wrote a re-implementation of much of
Numeric, called NumArray, to support their need to work on large memory-mapped
arrays, and arrays of mixed data type records \cite{STScI-slither}.  This
briefly caused the Numeric and NumArray communities to diverge, until 2005,
when Travis Oliphant embarked upon a major rewrite that aimed to be a ``best of
both worlds'' unification of Numeric and NumArray \cite{oliphant2006guide}.
This merger resulted in NumPy.  At the time, Oliphant was an assistant
professor of Electrical and Computer Engineering at Brigham Young University.

Over the next 10 years, NumPy attracted a relatively large pool of scientist
developers.  To benefit from these
user-developers, the project developed an increasingly strong culture of using
software-engineering practice to improve collaboration and reduce error
\cite{millman2014developing}. The NumPy team was early in adopting distributed
revision control and code review to improved collaboration on code, and
continuous testing so we run a large battery automated tests for every proposed
change in NumPy. At first, NumPy had rather sparse documentation, but the project
developed a new standard for ``docstrings''---text describing each NumPy
function---and used this standard to write comprehensive high-quality
documentation, integrated with the source code. The NumPy documentation format
quickly became a standard for the larger Python community.

@@

Hugunin identified the first factor in his initial description of the Numeric
package \cite{Hugunin-whitepaper}.  Numerical computing is usually one
component of a larger programming task.  Therefore:

@@

The second factor flows from Python's nature as an open-source language,
embedded within the open-source community.  Python attracts developers who like
to build and contribute.  As a result, NumPy has always been a library that was
developed by scientists in order to do their work.  Commercial specialist
numerical languages can encourage consumer-users, who pick up a tool and use
it, but do not expect to contribute to development.  In contrast, NumPy often
attracts user-developers, who want to work with NumPy, and are prepared to fix
and extend the library as they work. User-developers have been the fundamental
force in developing NumPy, and shaping it into a library that has been refined
in the furnace of practical scientific work.

@@

>>>>

NumPy is both a cause and an effect of a striking shift in the nature of
scientific computing.  In the traditional scientific software model, the user is a {\it consumer}.  Writing software is a task for specialists and
professionals, supported by income from selling the software to
scientists.  In the new model, the user is an {\it owner}.  The Python language, with
the NumPy library, provides a framework that scientists can consume,
but also encourages scientists to participate as programmers---to
write code for the library and the ecosystem of packages around it.  This has
fostered a new culture of contribution and collaboration in scientific
computing that resulted in a virtuous cycle of increasing use, and increasing
development.

@@

The first incarnation of NumPy was a package called Numeric, written in 1995
by Jim Hugunin, then completing his master's thesis at MIT on
superconductor-semiconductor junctions.  Hugunin based his package on previous
work by Jim Fulton, then working at the US Geological Survey, and acknowledges
many others in his initial report of the package \cite{Hugunin-whitepaper}.
Numeric provided an array object in Python, written in C, and linking to
standard fast implementations of linear algebra.
% https://stackoverflow.com/questions/26948776/where-did-the-term-broadcasting-come-from/26950256
% https://mail.python.org/pipermail/matrix-sig/1995-November/000143.html
% May want to mention Yorick here (origin of broadcasting idea)

As Hugunin points out, at the time he wrote Numeric, commercial languages for
interactive computing with numerical arrays were already well-established,
including Matlab and IDL.  Back then, it must have seemed a safe bet that
Numeric would not succeed in this market.  The commercial languages were
developed and maintained by large groups of full-time professional programmers,
supported by formal quality assurance, release managers, and writers of
scientific documentation.  In contrast, Numeric, and later NumPy, would
continue to be developed almost entirely by scientists, including graduate
students and faculty, without specific funding and often to the detriment of their careers.
% S: The "Back then" sentence sets up an expectation.  Later in this section, we then explain that Numeric/NumPy *was* successful, but we never explicitly close the loop: "despite the odds, great success".

@@

% S TODO: the first sentence below could flow a bit better
Around 1998, the Space Telescope Science Institute (STScI) software group began
to use Python heavily and, around 2000, wrote a re-implementation of much of
Numeric, called NumArray, to support their need to work on large memory-mapped
arrays and arrays of mixed data type records \cite{STScI-slither}.  This
briefly caused the Numeric and NumArray communities to diverge, until 2005,
when Travis Oliphant embarked upon a major rewrite that aimed to be a ``best of
both worlds'' unification of Numeric and NumArray \cite{oliphant2006guide}.
This merger resulted in NumPy.  At the time, Oliphant was an assistant
professor of Electrical and Computer Engineering at Brigham Young University.

Over the next 10 years, NumPy attracted a relatively large pool of scientist
developers.  To benefit from these
user-developers, the project developed an increasingly strong culture of using
software-engineering practice to improve collaboration and reduce error
\cite{millman2014developing}. The NumPy team was early in adopting distributed
revision control and code review to improve collaboration on code, and
continuous testing that runs a large battery of automated tests for every proposed
change to NumPy. At first, NumPy had rather sparse documentation, but the project
developed a new standard for ``docstrings''---text describing each NumPy
function---and used this standard to write comprehensive, high-quality
documentation, integrated with the source code\cite{vanderwalt2008scipy,harrington2008scipy,harrington2009scipy}. The NumPy documentation format
quickly became a standard for the larger Python community.

@@

Hugunin identified the first factor in his initial description of the Numeric
package \cite{Hugunin-whitepaper}.  Numerical computing is usually one
component of a larger programming task, therefore:

@@

The second factor flows from Python's nature as an open-source language,
embedded within the open-source community.  Python attracts developers who like
to build and contribute.  As a result, NumPy has always been a library that was
developed by scientists in order to do their work.  Commercial specialist
numerical languages can encourage consumer-users, who pick up a tool and use
it, but do not expect to contribute to development%<-sentence a bit tough to parse.
In contrast, NumPy often
attracts user-developers, who want to work with NumPy, and are prepared to fix
and extend the library as they work. User-developers have been the fundamental
force in developing NumPy, and shaping it into a library that has been refined
in the furnace of practical scientific work.

@@

% S: I think two issues are conflated here: the simple underlying memory model, and the API.  The API has developed organically, while the underlying memory model is a design decision that shapes much of how the rest of the library was built.

% S:
%
% Thinking about enabling factors, I'd characterize them into three categories:
%
% - Practical
% - Philosophical
% - Social
%
% E.g., practical: reason why we use Python (learn one language for everything); students don't have money, so want to avoid impracticalities of license dongles/servers, etc.
%
% Philosophical: science should be open, transparent; our software should be controlled by scientists, not designers that we don't have access to
%
% Social: joy of building these things together, friendly welcome into the community for many of us---appreciation of our work and hours
%

### 2020-01-06 summary.tex
Added notes and small edits

% - 2-3 sentences of basic-level introduction to the field;
Software libraries for array computation provide powerful notation
for accessing, manipulating, and operating on aggregate data in vectors,
matrices, or higher-dimensional arrays \cite{iverson1980notation}.
% - a brief account of the background and rationale of the work;
NumPy is the fundamental array computation library for the Python ecosystem
\cite{dubois2007guest,oliphant2007python,millman2011python,perez2011python}.
Not surprisingly, many important scientific research projects and results
crucially rely on NumPy.
It plays an essential role in many research analysis pipelines in fields as
diverse as physics, biology, astronomy, neuroscience, material science,
engineering, and chemistry.
In particular, NumPy played a fundamental role in discovering gravitational
waves\cite{pycbc} and the first imaging of a black hole\cite{eht-imaging}, and
lets us explore our universe\cite{jenness2018lsst}.
It is also one of the most commonly used machine learning tools among
enterprises \cite{451report2018}.
% - a statement of the main conclusions (introduced by the phrase 'Here we show' or its equivalent);
Here we provide an overview of the project and library organization,  highlight
some recent technical developments, and outline our future plans.
% - 2-3 sentences putting the main findings into general context so it is
%   clear how the results described in the paper have moved the field forwards.
NumPy continues to evolve to meet the needs of its community, to remain
relevant in an ever-changing computing landscape, and to be a fundamental
pillar of support for the next decade of data science in Python.

>>>>

% - 2-3 sentences of basic-level introduction to the field;
Software libraries for array computation provide powerful notation
for accessing, manipulating, and operating on data in vectors,
matrices, and higher-dimensional arrays \cite{iverson1980notation}.
% Stéfan: why "aggregate" data?
% - a brief account of the background and rationale of the work;
NumPy is the fundamental array computation library for the Python ecosystem
\cite{dubois2007guest,oliphant2007python,millman2011python,perez2011python}.
Not surprisingly, many important scientific research projects and results
depend on NumPy.
It plays an essential role in research analysis pipelines in fields as
diverse as physics, biology, astronomy, neuroscience, material science,
engineering, and chemistry.
For example, in astronomy, NumPy was fundamental to discovering gravitational
waves\cite{pycbc}, the first imaging of a black
hole\cite{eht-imaging}, and continues underpinning the future exploration of our universe\cite{jenness2018lsst}.
It is also one of the most commonly used machine learning tools among
enterprises \cite{451report2018}.
% - a statement of the main conclusions (introduced by the phrase 'Here we show' or its equivalent);
Here we provide an overview of the project and library organization, highlight
some recent technical developments, and outline our future plans.
% - 2-3 sentences putting the main findings into general context so it is
%   clear how the results described in the paper have moved the field forwards.
NumPy continues to evolve to meet the needs of its community, to remain
relevant in an ever-changing computing landscape, and to be a fundamental
pillar of support for the next decade of data science in Python.
